---
title: "Kryptos K2"
author: "Kenny Shirley"
date: "September 1, 2019"
output: 
  html_document:
    smart: FALSE
    toc: TRUE
    theme: readable
editor_options: 
  chunk_output_type: console
---
  
```{r global-options, include = FALSE}
## Global options
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE, 
                      cache = FALSE, 
                      comment = "")
```

```{r set-options, echo=FALSE}
#setwd("~/misc/kryptos")
options(width = 80)
library(dplyr,      quietly = TRUE)
library(knitr,      quietly = TRUE)
library(tidyr,      quietly = TRUE)
library(data.table, quietly = TRUE)
library(ggplot2,    quietly = TRUE)
library(plotly,     quietly = TRUE)
```


## Solving Kryptos 2 ("K2")

``` {r read-data}
cipher_text <- readLines("ciphertext.txt")
vigenere_table <- readLines("vigenere_table.txt")
```

For background on Kryptos, see [my introductory notebook here]().

Let's now solve K2, by stating a hypothesis, and testing it ("attacking" is a more fun word, though). See my K1 notebook for a description of how to decode a keyed Vigener cipher given a set of candidate keywords, and a bigram model for English.

We know that K2 consists of the 372 characters in lines 3 through 14 of the ciphertext half of the Kryptos sculpture.

``` {r K2}
K2 <- unlist(strsplit(paste(cipher_text[3:14], collapse = ""), ""))
```

Let's test the same hypothesis that we tested for K1:

Keyed Vigenere: keyword = single word in Scrabble list (n = 178,691); alphabet = "kryptos".

First, we'll read in some data and set up some necessary functions:

``` {r functions}
# function to decrypt a cipher using a vigenere table and a key:
decode_vigenere <- function(cipher, key, vigenere_table) {
  stopifnot(all(cipher %in% c(LETTERS, "?")))
  stopifnot(all(key %in% LETTERS))
  row_indices <- match(key, V[, 1])
  qmark <- cipher == "?"  # force a "?" to decode to itself
  r <- row_indices[0:(length(cipher[!qmark]) - 1) %% length(row_indices) + 1]
  D <- matrix(match(V[r, ], LETTERS) - rep(match(cipher[!qmark], LETTERS), 26), 
              nrow = length(cipher[!qmark]), 
              ncol = 26)
  # one entry in each row of D equals zero, except for question marks, which are rows of NA
  out <- rep("", length(cipher))
  out[!qmark] <- V[1, (which(t(D) == 0) - 1) %% 26 + 1]
  out[qmark] <- "?"
  out
}


# store the value of the Kryptos Vigenere table in a matrix:
V <- vector("list", 26)
alphabet <- rep(unlist(strsplit("KRYPTOSABCDEFGHIJLMNQUVWXZ", split = "")), 2)
for (i in 1:26) V[[i]] <- alphabet[1:26 + i - 1]
V <- do.call(cbind, V)

# read in the scrabble dictionary:
scrabble <- readLines("TWL06.txt")

# read in the 2-gram frequencies from the Google n-grams corpus:
two_gram <- fread("count_2l.txt", data.table = FALSE)
names(two_gram) <- c("letters", "frequency")

# normalize the frequencies to avoid integer overflow:
two_gram$normalized <- two_gram$frequency / min(two_gram$frequency)

# build the transition matrix for the bigram model of English:
two_mat <- matrix(0, 26, 26)
row_index <- match(substr(two_gram$letters, 1, 1), letters)
col_index <- match(substr(two_gram$letters, 2, 2), letters)
two_mat[cbind(row_index, col_index)] <- two_gram$normalized
rownames(two_mat) <- LETTERS
colnames(two_mat) <- LETTERS

# divide by row sums to get probabilities:
theta <- two_mat / rowSums(two_mat)

# get initial state distribution:
p <- rowSums(two_mat) / sum(two_mat)

# function to compute the log-likelihood of a particular snippet of text:
markov_score <- function(sequence, trans_matrix = theta, initial_probs = p) {
  n <- length(sequence)
  row_indices <- match(sequence[1:(n - 1)], LETTERS)
  col_indices <- match(sequence[2:n], LETTERS)
  as.numeric(log(p[match(sequence[1], LETTERS)])) + 
    sum(log(theta[cbind(row_indices, col_indices)]), na.rm = TRUE)
}
```

Next, we'll loop through all the possible keywords, computing the likelihood that each solution is a snippet of English language.

``` {r decode-vigenere-K2, echo = TRUE, eval = FALSE}
# Now, try all the words in the Scrabble dictionary for K2:
candidate <- vector("list", length(scrabble))
loglik <- numeric(length(scrabble))
for (i in 1:length(scrabble)) {
  candidate[[i]] <- decode_vigenere(cipher = K2,
                                    key = unlist(strsplit(scrabble[i], split = "")), 
                                    vigenere_table = V)
  loglik[i] <- markov_score(sequence = candidate[[i]])
}
```

``` {r save-results-K2, echo = FALSE, eval = FALSE}
saveRDS(candidate, file = "K2_candidate.rds")
saveRDS(loglik, file = "K2_loglik.rds")
```

``` {r load-results-K2, echo = FALSE, eval = TRUE}
candidate <- readRDS("K2_candidate.rds")
loglik <- readRDS("K2_loglik.rds")
```

``` {r show-K2-results}
# look at the top 10 most likely messages (and keywords):
top <- order(loglik, decreasing = TRUE)[1:10]

out <- data.frame(key = scrabble[top], 
                  loglik = loglik[top], 
                  message = sapply(candidate[top], paste, collapse = ""))

out
```

It worked again! The most likely solution is $e^{204}$ times more likely than the next most likely solution. The keyword is "ABSCISSA" and the message, with spaces and punctuation inserted, reads:

<pre>
IT WAS TOTALLY INVISIBLE HOWS THAT POSSIBLE? THEY USED THE EARTHS MAGNETIC FIELD X
THE INFORMATION WAS GATHERED AND TRANSMITTED UNDERGRUUND TO AN UNKNOWN LOCATION X 
DOES LANGLEY KNOW ABOUT THIS? THEY SHOULD ITS BURIED OUT THERE SOMEWHERE X
WHO KNOWS THE EXACT LOCATION? ONLY WW THIS WAS HIS LAST MESSAGE X 
THIRTY EIGHT DEGREES FIFTY SEVEN MINUTES SIX POINT FIVE SECONDS NORTH 
SEVENTY SEVEN DEGREES EIGHT MINUTES FORTY FOUR SECONDS WEST ID BY ROWS
</pre>

Recall that K2 had three question marks; here we used the rule that question marks decrypted to themselves. Also, note that the character "X" in the plaintext is used to separate sentences, although our bigram model for English was still interpreting it as an English character. This could be problematic for shorter messages. Our bigram model might reject a message with lots of "X" characters as unlikely to be English, but in fact, the "X" characters might just signify breaks between phrases or sentences. This is something to keep in mind for later.

One interesting aspect of K2 is that the above solution was accepted from roughly 1999 to 2006 (?), until Jim Sanborn revealed that there was a mistake in his original encoding. He claims to have omitted an "S" in the 24th position of the 14th row of the ciphertext. If an "S" is inserted here, the end of the message reads "... WEST X LAYER TWO" rather than "... WEST ID BY ROWS".

First, let's confirm that:

``` {r K2-version2}
K2_updated <- c(K2[1:364], "S", K2[365:372])
decode_vigenere(cipher = K2_updated,
                key = unlist(strsplit("ABSCISSA", split = "")), 
                vigenere_table = V)
```

Sure enough, inserting an "S" here decrypts to the new message, ending in "X LAYER TWO".


